How Significant was WW2 in the Development of the Computer?

### 2017-02-20 Draft 2/Section 1 - Lambda Calculus.tex
Insignificant tiny changes which I can't place
sec1: renamed R function to Y combinator

\subsubsection{Arithmetical definitions}

@@

\subsubsection{Logic and Conditions}

@@

With the fundamental boolean system of our computational system in place, we must next construct conditionals. The first, and simplest, of these will check if a number is equal to zero:
\begin{equation*}
\bm{\mathrm{Z}} = \lambda x.x\bm{\mathrm{F\lnot F}}
\end{equation*}
To understand this somewhat less than trivial concept, remember that \textbf{F} applied to any single argument yields the identity function $\lambda x.x$, and that \textbf{0} applied to any two arguments yields only the second:
\begin{equation*}
\bm{0}ab = (\lambda sz.z)ab = b
\end{equation*}
From this, it easier to see how the function works. Firstly with \textbf{0}:
\begin{gather*}
\begin{aligned}
\bm{\mathrm{Z0}} &= (\lambda x.x\bm{\mathrm{f\lnot F}})\bm{0}\\
&= \bm{\mathrm{0F\lnot F}}\\
&= \bm{\mathrm{\lnot F}}\\
&= \bm{\mathrm{T}}
\end{aligned}
\end{gather*}
Secondly with \textbf{2}:
\begin{gather*}
\begin{aligned}
\bm{\mathrm{Z2}} &= (\lambda x.x\bm{\mathrm{f\lnot F}})\bm{2}\\
&= \bm{\mathrm{2F\lnot F}}\\
&= (\bm{\mathrm{F}}(\bm{\mathrm{F}}\lnot)) \bm{\mathrm{F}}\\
&= (\bm{\mathrm{F}}(\lambda x.x))\bm{\mathrm{F}}\\
&= (\lambda x.x)\bm{\mathrm{F}}\\
&= \bm{\mathrm{F}} 
\end{aligned}
\end{gather*}

However, in order to develop other conditional tests we will need a 'predecessor' function - something surprisingly difficult under lambda calculus.
\subsubsection{Subtraction}

@@

Of course now the subtraction function can be found (using \textbf{D} for difference):
\begin{equation*}
\bm{\mathrm{D}} = \lambda ab.b\bm{\mathrm{P}}a
\end{equation*}
Bearing in mind the predecessor of \textbf{0} is \textbf{0}, this will produce \textbf{0} if $b>a$.
\subsubsection{Further Conditional Logic}

With this useful quirk of the \textbf{D} function, we can produce comparative logical operators:
\begin{equation*}
\begin{aligned}
\bm{\leq} &= \lambda ab.\bm{\mathrm{Z}}(\bm{\mathrm{D}}ab)
  & % seperation
\bm{\geq} &= \lambda ab.\bm{\mathrm{Z}}(\bm{\mathrm{D}}ba)
  \\
\bm{<} &= \lambda ab.\bm{\mathrm{Z}}(\bm{\mathrm{D}}a[\bm{\mathrm{S}}b])
  &
\bm{>} &= \lambda ab.\bm{\mathrm{Z}}(\bm{\mathrm{D}}[\bm{\mathrm{S}}b]a)
\end{aligned}
\end{equation*}
And through combination of these, we can create the equivalence and inequivalence operators (using $\bm{\doteq}$ for clarity from $=$:
\begin{equation*}
\begin{aligned}
\bm{\doteq} &= \lambda ab.\wedge(\bm{\leq}ab)(\bm{\geq}ab)\\
\bm{\neq} &= \lambda ab.\vee(\bm{<}ab)(\bm{>}ab)
\end{aligned}
\end{equation*}
\subsubsection{Recursion}

So far, Church's $\lambda$-Calculus is very close to the operation of a modern computation language. It has representation for boolean values, for conditional statements, for positive integers (signed integers can be expressed as a pair of a positive integer and a boolean for sign) and arithmetic, as well as set operations for booleans and comparators for integers - In fact, we are very close to our definition of computability save one thing. Continuity and repetition is one of the most significant aspects of computers, the power of this being the ability to transform a finite configuration into a potentially infinite result. This is done through one of the most remarkable functions in $\lambda$-Calculus, the recursor function:

\begin{equation*}
\bm{\mathrm{R}} = \lambda y.(\lambda x.y(xx))(\lambda x.y(xx))
\end{equation*}
\begin{equation*}
\begin{aligned}
\bm{\mathrm{R}}a &= (\lambda x.a(xx))[\lambda x.a(xx)]\\
&= a([\lambda x.a(xx)][\lambda x.a(xx)])\\
&= a(\bm{\mathrm{R}}a)
\end{aligned}
\end{equation*}
Therefore, the \textbf{R} function applied to any function $a$ applies the function $a$ to the function \textbf{R} applied to $a$ - if substitution continues, an infinite series of nested $a$ functions is created; an infinite result is produced from a finite configuration. Of course we can create finites as well by halting the function - an example of this would be the factorial function, where \textbf{!}a = \textbf{M}a(\textbf{!}[\textbf{P}a]) and \textbf{!0} = \textbf{1}. Firstly, a recursive definition:
\begin{equation*}
\begin{aligned}
\bm{\mathrm{\xi}} &= \lambda fa.\bm{\mathrm{Z}}a\bm{\mathrm{1}}[\bm{\mathrm{M}}a(f[\bm{\mathrm{P}}a])]\\
\bm{!} &= \bm{\xi}!
\end{aligned}
\end{equation*}
i.e. the factorial function is the $\bm{\xi}$ function with itself as the argument $f$. Unfortunately such a definition is illegal under $\lambda$-Calculus as strictly no functions have names - this is where the function \textbf{R} comes into play. To find the factorial of \textbf{3}:
\begin{equation*}
\begin{aligned}
\bm{\mathrm{R\xi 3}} &= \bm{\mathrm{\xi}}(\bm{\mathrm{R\xi}})\bm{\mathrm{3}}\\
&= \bm{\mathrm{Z31}}[\bm{\mathrm{M3}}([\bm{\mathrm{R\xi}}][\bm{\mathrm{P3}}])]\\
&= \bm{\mathrm{M3}}(\bm{\mathrm{R\xi}}\bm{2})\\
&= \bm{\mathrm{M3}}(\bm{\mathrm{M2}}(\bm{\mathrm{M11}}))\\
&= \bm{\mathrm{6}}
\end{aligned}
\end{equation*}
The recursive identity \textbf{R} allows for \textbf{R}$\bm{\xi}$ to be applied as the first argument to the $\bm{\xi}$ function - as a result, the procedure also processes \textbf{2} and \textbf{1}, before stopping at \textbf{0} (which produces simply \textbf{1}. To simplify, the factorial function:
\begin{equation*}
\bm{!} = \lambda x.\bm{\mathrm{R\xi}}x
\end{equation*}

>>>>

\subsection{Arithmetical definitions}

@@

\subsection{Boolean Logic}

@@

\subsection{Basic Comparison operators}

With the fundamental boolean system of our computational system in place, we must next construct comparison operators. The first, and simplest, of these will check if a number is equal to zero:
\begin{equation*}
\bm{\mathrm{Z}} = \lambda x.x\bm{\mathrm{F\lnot F}}
\end{equation*}
To understand this somewhat less than trivial concept, remember that \textbf{F} applied to any single argument yields the identity function $\lambda x.x$, and that \textbf{0} applied to any two arguments yields only the second:
\begin{equation*}
\bm{0}ab = (\lambda sz.z)ab = b
\end{equation*}
From this, it easier to see how the function works. Firstly with \textbf{0}:
\begin{gather*}
\begin{aligned}
\bm{\mathrm{Z0}} &= (\lambda x.x\bm{\mathrm{f\lnot F}})\bm{0}\\
&= \bm{\mathrm{0F\lnot F}}\\
&= \bm{\mathrm{\lnot F}}\\
&= \bm{\mathrm{T}}
\end{aligned}
\end{gather*}
Secondly with \textbf{2}:
\begin{gather*}
\begin{aligned}
\bm{\mathrm{Z2}} &= (\lambda x.x\bm{\mathrm{f\lnot F}})\bm{2}\\
&= \bm{\mathrm{2F\lnot F}}\\
&= (\bm{\mathrm{F}}(\bm{\mathrm{F}}\lnot)) \bm{\mathrm{F}}\\
&= (\bm{\mathrm{F}}(\lambda x.x))\bm{\mathrm{F}}\\
&= (\lambda x.x)\bm{\mathrm{F}}\\
&= \bm{\mathrm{F}} 
\end{aligned}
\end{gather*}

However, in order to develop other conditional tests we will need a 'predecessor' function - something surprisingly difficult under lambda calculus.
\subsection{Subtraction}

@@

Of course now the subtraction function can be found (using \textbf{D} for difference):
\begin{equation*}
\bm{\mathrm{D}} = \lambda ab.b\bm{\mathrm{P}}a
\end{equation*}
Bearing in mind the predecessor of \textbf{0} is \textbf{0}, this will produce \textbf{0} if $b>a$.
\subsection{Further Conditional Logic}

With this useful quirk of the \textbf{D} function, we can produce comparative logical operators:
\begin{equation*}
\begin{aligned}
\bm{\leq} &= \lambda ab.\bm{\mathrm{Z}}(\bm{\mathrm{D}}ab)
  & % seperation
\bm{\geq} &= \lambda ab.\bm{\mathrm{Z}}(\bm{\mathrm{D}}ba)
  \\
\bm{<} &= \lambda ab.\bm{\mathrm{Z}}(\bm{\mathrm{D}}a[\bm{\mathrm{S}}b])
  &
\bm{>} &= \lambda ab.\bm{\mathrm{Z}}(\bm{\mathrm{D}}[\bm{\mathrm{S}}b]a)
\end{aligned}
\end{equation*}
And through combination of these, we can create the equivalence and inequivalence operators (using $\bm{\doteq}$ for clarity from $=$:
\begin{equation*}
\begin{aligned}
\bm{\doteq} &= \lambda ab.\wedge(\bm{\leq}ab)(\bm{\geq}ab)\\
\bm{\neq} &= \lambda ab.\vee(\bm{<}ab)(\bm{>}ab)
\end{aligned}
\end{equation*}
\subsection{Recursion}

So far, Church's $\lambda$-Calculus is very close to the operation of a modern computation language. It has representation for boolean values, for conditional statements, for positive integers (signed integers can be expressed as a pair of a positive integer and a boolean for sign) and arithmetic, as well as set operations for booleans and comparators for integers - In fact, we are very close to our definition of computability save one thing. Continuity and repetition is one of the most significant aspects of computers, the power of this being the ability to transform a finite configuration into a potentially infinite result. This is done through one of the most remarkable functions in $\lambda$-Calculus, the \textbf{Y} combinator:

\begin{equation*}
\bm{\mathrm{Y}} = \lambda y.(\lambda x.y(xx))(\lambda x.y(xx))
\end{equation*}
\begin{equation*}
\begin{aligned}
\bm{\mathrm{Y}}a &= (\lambda x.a(xx))[\lambda x.a(xx)]\\
&= a([\lambda x.a(xx)][\lambda x.a(xx)])\\
&= a(\bm{\mathrm{Y}}a)
\end{aligned}
\end{equation*}
Therefore, the \textbf{Y} function applied to any function $a$ applies the function $a$ to the function \textbf{Y} applied to $a$ - if substitution continues, an infinite series of nested $a$ functions is created; an infinite result is produced from a finite configuration. Of course we can create finites as well by halting the function - an example of this would be the factorial function, where \textbf{!}a = \textbf{M}a(\textbf{!}[\textbf{P}a]) and \textbf{!0} = \textbf{1}. Firstly, a recursive definition:
\begin{equation*}
\begin{aligned}
\bm{\mathrm{\xi}} &= \lambda fa.\bm{\mathrm{Z}}a\bm{\mathrm{1}}[\bm{\mathrm{M}}a(f[\bm{\mathrm{P}}a])]\\
\bm{!} &= \bm{\xi}!
\end{aligned}
\end{equation*}
i.e. the factorial function is the $\bm{\xi}$ function with itself as the argument $f$. Unfortunately such a definition is illegal under $\lambda$-Calculus as strictly no functions have names - this is where the \textbf{Y} combinator comes into play. To find the factorial of \textbf{3}:
\begin{equation*}
\begin{aligned}
\bm{\mathrm{Y\xi 3}} &= \bm{\mathrm{\xi}}(\bm{\mathrm{Y\xi}})\bm{\mathrm{3}}\\
&= \bm{\mathrm{Z31}}[\bm{\mathrm{M3}}([\bm{\mathrm{Y\xi}}][\bm{\mathrm{P3}}])]\\
&= \bm{\mathrm{M3}}(\bm{\mathrm{Y\xi}}\bm{2})\\
&= \bm{\mathrm{M3}}(\bm{\mathrm{M2}}(\bm{\mathrm{M11}}))\\
&= \bm{\mathrm{6}}
\end{aligned}
\end{equation*}
The \textbf{Y} combinator allows for \textbf{Y}$\bm{\xi}$ to be applied as the first argument to the $\bm{\xi}$ function - as a result, the procedure also processes \textbf{2} and \textbf{1}, before stopping at \textbf{0} (which produces simply \textbf{1}. To simplify, the factorial function:
\begin{equation*}
\bm{!} = \lambda x.\bm{\mathrm{Y\xi}}x
\end{equation*}
